\mychapter{The Whole of the Holy Life}{Ajahn Karuṇadhammo}{June 2013}

With many people away, it's so quiet I can hear the water dripping over 
the sound of our breathing. We can have our own little retreat here 
today, which is the way it should be all the time. We can use every 
opportunity to practice mindfulness---to be aware of where our body is 
and what our mind is doing and ask ourselves, \emph{Is this a skillful 
state of mind or is it an unskillful state of mind? Am I distracted? Am 
I in touch? Am I so absorbed in what I'm doing that I am not really 
seeing what's going on with the people around me?}

The title of Luang Por Pasanno's daylong retreat at Spirit Rock today 
is ``The Whole of the Path: the Fruits of Spiritual Friendship.'' Many 
people know the sutta where Ānanda says to the Buddha, ``The Saṅgha 
is half of the holy life,'' and the Buddha replies, ``Don't say that 
Ānanda; it's the whole of the holy life.'' The opportunity to practice 
with people is one of the unique features of our community. Where would 
we be if we didn't have support from people when we are having 
difficulties or the feedback from people when we are missing the mark? 
The qualities of mindfulness, self-awareness, and sensitivity to other 
people are strengthened with the reflective guidance of spiritual 
friendship. Many of us are in need of individualized guidance for our 
training, and we support each other in this. We embody the sense of 
community and the sense of care and concern for each other. This is 
where the fruits of living in a community can truly be seen, especially 
in juxtaposition to people who lack support and have difficulties going 
it alone.

It can sometimes be challenging to rejoice in and take advantage of the 
opportunity to practice within a community. Because of this, it is 
useful to remind ourselves that it may not be here forever. 
Circumstances change, communities change, and we never know where we 
are going to find ourselves next. If we keep this in mind, we can use 
the opportunity we have here to be sensitive to each other, mindful 
around one another, orient ourselves to our present experience and feel 
a sense of appreciation for the community we live in.

