\mychapter{Eye of the Hurricane}{Ajahn Karuṇadhammo}{November 2013}

We have a full community once again. With all of the comings and 
goings, right now there seems to be a confluence of comings---everybody 
is present within the community for at least a little while. We also 
have a fairly sizable group of long-term residents, visitors, and 
guests---more people than we've been accustomed to in the past few 
weeks. Notice the effect of that. Try to keep in mind the sense of 
being in a community and learning from all the experiences of living 
with and relating to different people. We work together, make 
decisions, offer suggestions, receive suggestions, and experience the 
whole interplay of human interaction. We can reflect on how to do this 
skillfully and remember the importance of staying centered while in 
activity, trying not to let the senses be pulled out in different 
directions. It helps to take a moment and come back to the center of 
awareness in the body.

Many of us sometimes have a fantasy of being in total solitude and not 
having to interact with other people. I know I sure do. Or we imagine 
living in a small community with only one or two other people. However, 
one of the advantages of living and working together in a sizable 
community with a fair amount of engagement is the quality of learning 
that takes place and the understanding we develop when we stay in the 
center of awareness without being pulled out into many different 
directions. It's like being in the eye of a hurricane where there are 
very strong winds around us---winds that could cause anxiety, fear, 
disturbance, or agitation. But right in the middle of the storm is the 
eye where it is very calm and still. We can place ourselves in that 
center so that we are alert and aware, where our senses are restrained 
and not moving out. Within this center, there is a receptivity to 
what's going on so that we are not totally isolated from the 
environment. We can respond appropriately, but the attention stays 
focused, particularly on the body. The body is the best place for 
maintaining that sense of presence. So as we move through the day we 
can try to stay in the eye of the hurricane.

