\mychapter{Looking Back on the Effort}{Ajahn Yatiko}{July 2012}

Yesterday, Luang Por Sopah was giving some reflections about Wat Pa 
Pong and what it was like in the early days. While Luang Por spoke, I 
looked over at Ajahn Pasanno and saw him smiling brightly as he 
listened. I could almost see the wheels in Ajahn Pasanno's head turning 
as his memories were being churned up. It seemed to me that he might 
have been thinking, ``I know what Luang Por Sopah is talking about, and 
no one else here can really know what it was like living with Luang Por 
Chah. But I do know and I do remember.''

I recall somebody saying once that after years of difficult practice 
have passed and time has put some perspective on one's life, then one 
can look back on all the effort and suffering involved in the practice 
with real appreciation, delight, and gratitude. This is very different 
from the suffering that arises when one becomes obsessed with personal 
ambitions, worries, and concerns. When we look back on those sorts of 
past experiences, there tends to be a feeling of loss and a sense of 
time wasted.

But again, when we look back at the effort we have put into our 
practice, how we have endured and cultivated the path, the beautiful 
teachings that we loved, respected, imbibed, and listened to, then the 
heart feels bright, warm, soft, and rich. That's something that I 
thought I clearly saw in Ajahn Pasanno yesterday as Luang Por Sopah was 
recollecting his life at Wat Pa Pong.

We can loose touch with the inspiring memories of our wholesome efforts 
due to the immediacy of the daily difficulties, frustrations, and 
temporary setbacks we experience. And with these daily concerns, we can 
easily forget that we're engaged in something very beautiful and noble, 
so that, from time to time, we need uplifting, we need to refresh 
ourselves. That's where these inspiring memories of our past efforts 
come in. They can uplift and refresh our practice, and so are worth 
bringing up, reflecting upon, and appreciating. They remind us that we 
have been engaged in something very beautiful and noble, indeed.

