\mychapter{A Positive Encounter With Death}{Ajahn Yatiko}{December 2012}

There are many aspects of life that we bring to mind, consciously and 
repetitively, as part of our training. One focus of our contemplation 
is on the nature of mortality, the fact that we're all going to die.

I recently had quite a positive experience in this regard. Last night 
there was a very strong and sharp pain in my chest that lasted several 
hours. It was running down my arm and the side of my neck and seemed 
very similar to a heart attack. But I was quite clear that it wasn't a 
heart attack because the pain felt slightly off to the right rather 
than to the left where the heart is. During this experience, I found 
that I could easily pretend and convince myself for a moment that this 
was a real heart attack.

Physically, it was very painful and difficult to breathe or walk, and I 
had to sit down for several hours. Interestingly, the earlier part of 
the day had been quite challenging, so that in comparison with that, it 
actually felt good to suddenly be face-to-face with death. Everything 
else fell away completely. All the concerns of the day, in a flash, 
disappeared from my radar.

Two experiences came up quite naturally from this episode. One was a 
recollection that my life as a monk has been pretty good. Certainly 
there have been ups and downs, successes and failures, morally and 
otherwise. But on the whole, I felt ready to die. I thought, 
\emph{That's okay}. The other experience was having mindfulness 
throughout the entire episode. I felt that I had stepped outside the 
body and was observing the situation from above, contemplating what was 
happening. I was able to see that there was this human body and this 
process of attachment and reflection. It was a bit surprising. I was 
looking at the whole situation from this perspective of not-self, and 
was mindful of that as well.

It was comforting for me to know that it was possible to have this 
response to the perception of death. Everything that seemed so heavy 
and weighed me down throughout the day suddenly disappeared like fog on 
a sunny day. It just vanished.

