\mychaptertoc{The Present Moment and the Illusion of Time}{The Present
Moment and\\the Illusion of Time}
{Ajahn Yatiko}{September 2012}

When undertaking the life of a monastic, we have a direction and a goal 
in mind. There's something we're aiming for, and it's important to 
reflect on that with some frequency. From time to time we can ask 
ourselves, \emph{What's the direction of my life? What's the goal of my 
life?}

We can also reflect on where this goal is located. With understanding 
we can see that it is located in the present moment. It's right now. 
Many of us live our lives thinking about the future most of the time. 
As we get older, we may think more and more about the past. But either 
way, we're not in the present moment. This is particularly apparent in 
our meditation, where the work is to bring our minds into the present 
and cut through the delusions we create around notions of time. The 
entire universe we attach to---what we like, dislike, dread, hate, 
fear, love---is a creation of our minds. We've created it in the 
present moment. Then we create this illusion of time and a whole 
universe of \emph{saṃsāra}, which we feel trapped in. But the whole 
creation of time is something we do now, in the present moment.

Bringing the mind to the present moment is hard work, because, over the 
years, we have pursued the habit of creating time, so that now this 
habit has great momentum and it keeps pushing us into the past and 
future. As a result, we could say, it takes a real `balancing act' to 
overcome that momentum---to bring ourselves back to a place that is 
centered and balanced in the present moment. Doing this is a matter of 
balance rather than effort.

When we see clearly, we clearly see that there's only the present 
moment. That's all there has ever been. It takes commitment and 
determination to see this and take it seriously. There is no worldly 
benefit to be gained from returning to the present. No one knows if we 
are in the present moment or not. No one will praise us for it, and we 
won't get any money for it---we don't get anything worldly out of it at 
all. But by continuing to return the mind to the present moment, we can 
eliminate having to bear the weight of the whole saṃsāric universe 
on our shoulders. No wonder coming back to the present is such a relief.

