\mychaptertoc{A Foundation of Love and Acceptance}
{A Foundation of Love\\and Acceptance}
{Ajahn Yatiko}{August 2013}

In this community at Abhayagiri, we need to take care of each other. If 
we don't have a foundation of love and care for each other, then the 
foundation of our lives is on very shaky ground.

I have a tremendous amount of respect and affection for each person 
here, no matter what their personality traits may be. It is important, 
I believe, that we give each other the space needed to be who we are. 
We conform externally; we have these robes, shave our heads, and look 
similar. Yet we are all completely different individuals, remarkably 
so. We need to allow everybody to be who and what they are---to be 
themselves. We mustn't start from a position of being unable to accept 
something or someone. The starting point in practice is being able to 
accept other people and to accept ourselves.

With that as a foundation, we then work with the defilements. It is not 
as if we work with the defilements as a condition for accepting 
ourselves---\emph{Once my defilements are under control, then I'll 
accept myself.} No. We start with acceptance, and from there we can 
work with whatever difficulties arise. Likewise, there's no need to 
expect other people to live up to our standards of monastery etiquette, 
harmony, or even our standards for being happy. It is okay to have a 
miserable day, to be gloomy for quite awhile. We are what we are, and 
we can accept each other whether we are happy, sad, anxious, excited, 
or afraid. That is a very good starting position on which we can build 
our practice.

