\mychapter{Appreciating the Goodness of Others}{Luang Por Pasanno}{June 
2013}

There are lots of us here living together in community, sharing the 
same space. So it's vital that we share the space harmoniously to 
ensure that things get done, like the external work of taking care of 
the monastery and the internal work of spiritual practice. To lay a 
foundation for living together harmoniously, there is a need to develop 
a sense of \emph{kataññū}, which is usually translated as 
``gratitude.'' Now the word \emph{gratitude} is a bit loaded in the 
West, sort of like somebody is standing behind us waving his finger 
saying, ``You should be grateful.'' That's quite a repelling image, and 
it doesn't convey the real essence of kataññū, which is the faculty 
within us that recognizes and appreciates the goodness of others.

It's important to exercise this faculty of kataññū to counteract the 
mind's tendency to focus on the things that irritate us, the flaws, the 
unskillful tendencies. That's where our minds go, and we tend to weigh 
ourselves down with negativity. I'm not suggesting we ignore or gloss 
over the shortcomings we see. Rather, this is about moving away from 
pointless negativity and, instead, recognizing and paying attention to 
the things people do that are skillful and wholesome.

Living in a monastery as we do, it's easy to take for granted the 
fundamental goodness of others. Goodness can seem so ordinary to 
us---it's part of the culture we live in here, and the standards we 
keep in that regard are quite high. So we may need to make a deliberate 
effort to keep recognizing this ``ordinary'' goodness and to engender a 
sense of gratitude for its presence. We make the effort to do this, 
because this gratitude nourishes our ability to live skillfully, and it 
helps establish wholesome states of mind. With our meditation, it's 
much easier to become peaceful when this quality of gratitude is 
present in the mind as opposed to when we experience negativity, which 
turns the mind toward all the flaws we perceive in everyone. So pay 
attention to gratitude---kataññū---learn to appreciate the goodness 
of others, which will have the most beneficial results for our practice 
and well-being.

