\mychaptertoc{Becoming a Somebody Forgetting About Everybody}
{Becoming a Somebody\\Forgetting About Everybody}
{Luang Por Pasanno}{April 2013}

As human beings, we each have a propensity for wanting to become a 
distinct \emph{somebody}. This propensity shows itself almost 
constantly, both in formal practice and in our ordinary, day-to-day 
activities. It has to do with the hopeless search for worldly security 
and stability. Most people spend their whole lives on that search, but 
instead of finding security and stability, what they get is a sense of 
isolation---not really attuning to themselves or to the world around 
them. If we're not careful, this can be a danger for us here, as well.

A prime example is the way we relate to our assigned duties and chores. 
We become the somebody who manages the book distribution, or the 
somebody who looks after the computers, or the somebody who does the 
trash, or whatever. We home in on the particular thing we're doing, and 
then conjure up a sense of ourselves as being the somebody who does 
that thing. Then it's easy to think: \emph{Alright, I have my task. I 
don't have to do anything more than that. I don't have to be concerned 
anymore with what needs to be done outside of my job. I'm safe and 
secure; things are okay.}

That sort of thinking blocks us from developing mindfulness and 
attentiveness in huge areas of our lives---the world we're in and 
what's happening around us. Thinking like that can cause us to feel 
isolated. We set ourselves apart from the community, weakening our 
connection with all the other somebodies here who are doing other 
things that need to be done.

To avoid these dangers, rather than being somebodies who do only what 
we're assigned, we can instead be attentive to whatever might need 
doing. We can ask ourselves: \emph{How can I help? How can I support 
the other people around me? What would be of service?} When everyone 
reflects in that way, and acts accordingly, the whole community can 
enjoy a sense of well-being and happiness. And we may personally learn 
that there's no real need to make ourselves into somebodies at all.

