\mychapter{The Natural Result}{Luang Por Pasanno}{July 2012}

This morning we chanted the Dhammacakkappavattana Sutta in English. 
There's always something that strikes me in that sutta, both hearing it 
and reflecting on it. For instance, the Buddha says the Middle Way and 
the Eightfold Path give knowledge and understanding and lead to peace 
and awakening. That's the function of the path. When we are aligned 
with the Middle Way, we are in balance---this happens naturally.

In \emph{Pāḷi}, the word \emph{karaṇī,} as in 
\emph{cakkhukaraṇī,} \emph{ñāṇakaraṇī}, means ``to make'' or 
``to produce.'' It's an emphatic verb. This suggests that when we put 
our practice into balance, a good result will be produced. It's helpful 
to develop a sense of confidence in this natural cause-and-effect 
relationship.

How do we achieve this balance? Again, we align ourselves with the 
Middle Way, which means not getting caught by the extremes of either 
sensual indulgence or ascetic deprivation; not seeking delight---always 
looking for something to gratify ourselves---and not pushing things 
away, trying to get rid of them.

Ajahn Chah used to say that there are basically two desires in the 
mind: one is the desire to get something, which has the energy of 
moving toward; the other is the desire to get rid something, not 
wanting to deal with it, which has the energy of pushing away. We see 
these two desires at work all the time---as we perform our tasks and 
duties, while engaging with people, and when we're off on our own. The 
mind tends to lean toward these two desires. It's not in balance; it's 
not following the Middle Way. But when we come back to a place of being 
attentive to balance, attentive to the Middle Way, bringing forth the 
factors of the path, then the natural results are knowledge and 
understanding and a sense of peace.

As we go about our daily chores, our day-to-day living in the 
monastery, we can make this simple effort of looking after the path, 
looking after those aspects of training the Buddha laid down for us. 
When we do this, we will produce the natural results of the Middle Way.

