\mychaptertoc{The Dhamma of the Buddhas Is Everywhere}
{The Dhamma of the Buddhas\\Is Everywhere}
{Luang Por Pasanno}{May 2012}

As we engage in activity, we can ground our awareness in the body and 
use it as a helpful anchor for our practice. We do this by having a 
tactile sense of contact with the ground, contact with physical 
movement and contact with the material world around us. This brings 
awareness into the body, but it's also important how this awareness is 
held. The Buddha's way of teaching and training is through questioning 
and reflection. We're not trying to apply a catechism of, ``The body is 
like this,'' repeating it over and over again. The awareness is an 
exploration, an investigation. We are asking ourselves, \emph{What is 
the experience I'm having? Does it feel comfortable or not? Does it 
feel wholesome or not? Is this what mindfulness is, or not?} This kind 
of questioning is not an intellectual analysis; rather, we are trying 
to use contact as a mirror for reflecting and seeing more clearly.

We should, of course, reflect on the mind as well. \emph{What's my 
mental state right now? What's the intention? What's the volition 
within the mind? Is this skillful or not skillful? How do I create and 
maintain a quality of skillfulness?} This is the kind of investigation 
we can carry with us throughout the day. Whether we are sitting in the 
Dhamma Hall, back at our \emph{kuṭis}, or doing our chores, 
reflecting in this way internalizes our experience. It gives us an 
anchor for recognizing whether we're on track or not, whether we're 
according with Dhamma or have lost the plot.

We need to develop these essential skills so that they permeate every 
aspect of our practice, no matter where we are or what we are doing. As 
Ajahn Chah and so many other teachers have often said, ``The Dhamma of 
the Buddhas is everywhere. We don't have to go to the Pali Canon to 
find it. We don't have to be sitting in a monastery to find it.'' It's 
the ability we have to recognize and see the truth, to see the way 
things truly are. It's everywhere. We can connect with it, we can 
contact it and we can experience it.

