\mychapter{The Spider in Its Web}{Luang Por Pasanno}{October 2013}

When the mind seeks objects for sense contact, it almost always finds 
the objects it seeks. When it seeks objects of stimulation, interest, 
gratification, aversion, or irritation, it can easily find them, 
because there's always something around to be excited or irritated 
about. That's why the quality of sense restraint is so important. We 
exercise this quality by paying close attention to the process of sense 
contact, and by learning to experience that contact---sights, sounds, 
smells, tastes, touch, and mental objects---without getting hooked into 
the desire and aversion, the liking and disliking, that tends to happen 
when sense restraint is absent.

Ajahn Chah used the image of a spider sitting in the center of its web, 
alert and waiting. As soon as an insect touches the web, the spider 
goes out, grabs the insect, wraps it up, and brings it back to the 
web's center for eating. In a similar way, we have a web made from our 
senses, and at its center is our heart and our ability to be aware. 
When this web is touched by some form of contact, we can draw that into 
the heart and---exercising sense restraint---we can reflect, 
investigate, and deal with that contact skillfully, according to 
Dhamma, without getting swept up in reaction and proliferation. When we 
take this quality of sense restraint as our center, we carry it into 
our daily lives of social interaction, duties and chores, and into 
formal meditation as well.

During the morning work period and at mealtime, if we let the mind get 
swept up into reactions and proliferations arising from social contact, 
when we go back to our dwellings in the afternoon, the mind will be 
chewing on it for perhaps hours, so that meditation becomes 
problematic. But if we are well established in sense restraint during 
such times of social contact, afterward, the mind will be steady and 
stable---it will be in a state much more conducive to having a fruitful 
meditation. As we endeavor to establish continuity in our practice, we 
recognize sense restraint as a key aspect of that. Establishing a 
continuity of sense restraint ensures that the food in our web will be 
nourishing.

