``Beginning our day \ldots{} ''

These quintessential words are spoken by Luang Por Pasanno before he 
begins each of his morning reflections.

Five days a week, at Abhayagiri's morning meeting, work tasks are 
assigned to the residents and guests living in the monastery. Shortly 
thereafter, one of the senior monks offers a brief Dhamma reflection so 
that the residents and guests have something meaningful to recollect 
throughout their day.

These talks are given spontaneously and often address an event that is 
about to occur, a condition that is already present in the monastery, 
or a general teaching on Dhamma. The most common thread through all the 
reflections is that of practicality: distilling the most important 
teachings of the Buddha into pertinent and applicable practices. Though 
many different teachings are touched upon, the fundamental aim is to 
encourage the abandonment of the unwholesome, the cultivation of the 
wholesome, and the purification of the mind.

While several of these teachings may be read together at one time, 
readers might find it more useful to focus on a single reflection so 
they can easily recollect, contemplate, and make use of it throughout 
their day.

This book was made possible through the contributions of many people. 
Over ten years ago, Pamela Kirby initiated the project when she placed 
a recorder in front of one of the senior monks during a morning 
reflection and proposed that a book be written. Matthew Grad, Jeff 
Miller, Ila Lewis, Ray Peterson, and Laurent Palmatier were the main 
substantive editors of the material, enduring the long and difficult 
process of editing the transcripts into compact and well-written 
teachings. Ruby Grad and Pamela Kirby completed the copy editing. 
Shirley Johannesen compiled the glossary. David Burrowes, Dee Cope, 
Josh Himmelfarb, Evan Hirsch, Jeanie Daskais, Anag캐rika John 
Nishinaga, and members of the Lotus Volunteer Group: Viveka and Wendy 
Parker all helped with further refining of the text. Sumi Shin designed 
the cover. Jonathan Payne took the cover photos. Michael Smith tumbled 
the stones for the back cover.

For several years, Khemako Bhikkhu recorded the senior monks' 
reflections. Kovilo Bhikkhu and Pesalo Bhikkhu provided corrections on 
an early draft of the book. Suhajjo Bhikkhu generously dedicated a 
significant amount of time on the overall book design and typesetting 
of the text.

The Kata침침ut캐 group of Malaysia, Singapore, and Australia generously 
brought this book into full production.

May these teachings bring insight into the nature of Dhamma and provide 
a pathway toward the development of true peace and contentment.

Cunda Bhikkhu\\
Abhayagiri Monastery,\\
Redwood Valley, California\\
May 15th, 2014

'Preface'
'\n'
u'\u201cBeginning our day ... \u201d'
'\n'
'These quintessential words are spoken by Luang Por Pasanno before he begins each of his morning reflections.'
'\n'
u'Five days a week, at Abhayagiri\u2019s morning meeting, work tasks are assigned to the residents and guests living in the monastery. Shortly thereafter, one of the senior monks offers a brief Dhamma reflection so that the residents and guests have something meaningful to recollect throughout their day.'
'\n'
'These talks are given spontaneously and often address an event that is about to occur, a condition that is already present in the monastery, or a general teaching on Dhamma. The most common thread through all the reflections is that of practicality: distilling the most important teachings of the Buddha into pertinent and applicable practices. Though many different teachings are touched upon, the fundamental aim is to encourage the abandonment of the unwholesome, the cultivation of the wholesome, and the purification of the mind.'
'\n'
'While several of these teachings may be read together at one time, readers might find it more useful to focus on a single reflection so they can easily recollect, contemplate, and make use of it throughout their day.'
'\n'
u'This book was made possible through the contributions of many people. Over ten years ago, Pamela Kirby initiated the project when she placed a recorder in front of one of the senior monks during a morning reflection and proposed that a book be written. Matthew Grad, Jeff Miller, Ila Lewis, Ray Peterson, and Laurent Palmatier were the main substantive editors of the material, enduring the long and difficult process of editing the transcripts into compact and well-written teachings. Ruby Grad and Pamela Kirby completed the copy editing. Shirley Johannesen compiled the glossary. David Burrowes, Dee Cope, Josh Himmelfarb, Evan Hirsch, Jeanie Daskais, Anag\u0101rika John Nishinaga, and members of the Lotus Volunteer Group: Viveka and Wendy Parker all helped with further refining of the text. Sumi Shin designed the cover. Jonathan Payne took the cover photos. Michael Smith tumbled the stones for the back cover.'
'\n'
u'For several years, Khemako Bhikkhu recorded the senior monks\u2019 reflections. Kovilo Bhikkhu and Pesalo Bhikkhu provided corrections on an early draft of the book. Suhajjo Bhikkhu generously dedicated a significant amount of time on the overall book design and typesetting of the text.'
'\n'
u'The Kata\xf1\xf1ut\u0101 group of Malaysia, Singapore, and Australia generously brought this book into full production.'
'\n'
'May these teachings bring insight into the nature of Dhamma and provide a pathway toward the development of true peace and contentment.'
'\n'
'Cunda Bhikkhu'
'Abhayagiri Monastery,'
'Redwood Valley, California'
'May 15'
'th'
', 2014'
